---
title: "VTPEH 6105 - Homework 04"
author: "Ruiting Zhu"
date: "2026-2-19"
editor_options:
  chunk_output_type: console
output:
  pdf_document:
    toc: true
    toc_depth: 1
    number_sections: true
    latex_engine: xelatex

---

```{r setup, include = FALSE}

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

```

# Instructions

This homework must be submitted on Canvas as a **R markdown-generated PDF** file (from the .Rmd template available on the Assignment page), **together with the source code** (the .rmd file). Submissions in any other format will be considered late until the requested format has been submitted; failure to submit the requested format within 10 days will lead to a 0 to the assignment.

Questions should be answered with **literate, meaningful answers** when relevant. Read all the questions before answering to ensure a good flow of your answers (e.g., no redundancy).

This homework aims at helping you:

1. Practice data analysis and hypothesis testing.
2. Practice peer-reviewing and critical assessment of published literature. 

# Peer-review 

## Introduction

*This part of the homework will focus on the review of Gold, C.H., Malmberg, B., McClearn, G.E., Pedersen, N.L. and Berg, S., 2002. Gender and health: a study of older unlike-sex twins.* The Journals of Gerontology Series B: Psychological Sciences and Social Sciences*, 57, 168-176. [https://doi.org/10.1093/geronb/57.3.s168](https://doi.org/10.1093/geronb/57.3.s168).* ***Read the full article*** *before answering the questions.* 

## Guiding questions

### Study overview

#### What is the overall research question? Answer in 1-3 sentences.

The study examines whether there are gender differences in health conditions, symptoms, and self-rated health among older adults by comparing brothers and sisters in unlike-sex twin pairs. The goal is to determine whether men and women differ in health outcomes when genetic and early-life environmental factors are controlled.

#### Briefly describe the study design. Answer in 1-3 sentences.

This is a population-based observational study using survey data from 605 pairs of unlike-sex twins in Sweden. The researchers used a matched-pair design and compared health outcomes between sisters and brothers using paired statistical tests.

### Methods

#### Which statistical test is used for the analyses?

Paired-sample t-tests were used to compare health measures between male and female twins.

#### What assumptions do the data need to verify for this test to be applicable?

For a paired t-test: The dependent variable is continuous. The observations are paired/matched. The differences between pairs are approximately normally distributed. Independence between pairs.
#### Does this publication report sufficient information to assess whether those assumptions are verified? Indicate which elements in the text, tables, or figures allow you to assess whether the assumptions are verified – if one or several are missing, indicate how this could be fixed by the authors of the study.

The publication does not provide sufficient information to fully assess whether the assumptions of the paired t-test are met. Although the authors report means, standard deviations, and sample sizes, they do not present any diagnostic plots or statistical tests to evaluate the normality of the paired differences. In addition, there is no discussion of potential outliers. To improve the assessment of assumptions, the authors could include histograms, Q–Q plots, or results of normality tests for the paired differences, as well as sensitivity analyses using non-parametric methods.

### Focusing on the last line of Table 3

#### What are the 2 variables of interest? Indicate what type of variable they are.

The two variables of interest are gender (male vs. female, categorical variable) and the total number of cardiovascular conditions (continuous variable).

#### What are the null and alternative hypotheses?

Null hypothesis (H₀): There is no difference in the mean number of total cardiovascular conditions between male and female twins.

Alternative hypothesis (H₁): There is a difference in the mean number of total cardiovascular conditions between male and female twins.

#### What are the two point estimates we aim to compare?

The two point estimates are the mean number of total cardiovascular conditions among female twins and the mean number among male twins.

#### How do you interpret the results? Explain the whole process (assessment of the null hypothesis and interpretation) and compare your interpretation to the authors’ interpretation presented in the Discussion section.

According to Table 3, the mean number of total cardiovascular conditions is 0.83 for females and 0.84 for males, with a paired t-test value of −0.11 and a non-significant p-value. Since the p-value is greater than 0.05, we fail to reject the null hypothesis. This indicates that there is no statistically significant difference in total cardiovascular conditions between male and female twins.

The authors also conclude in the Discussion section that there are no gender differences in total cardiovascular conditions. Therefore, this interpretation is consistent with the authors’ conclusions.

### Focusing on the first line of Table 4

#### What are the 2 variables of interest? Indicate what type of variable they are.

The two variables of interest are gender (male vs. female, categorical variable) and the number of physical symptoms reported (continuous variable).

#### What are the null and alternative hypotheses?

Null hypothesis (H₀): There is no difference in the mean number of physical symptoms between male and female twins.

Alternative hypothesis (H₁): There is a difference in the mean number of physical symptoms between male and female twins.

#### What are the two point estimates we aim to compare?

The two point estimates are the mean number of physical symptoms reported by female twins and the mean number reported by male twins.

#### How do you interpret the results? Explain the whole process (assessment of the null hypothesis and interpretation) and compare your interpretation to the authors’ interpretation presented in the Discussion section.

Table 4 shows that females reported an average of 3.21 physical symptoms, while males reported an average of 2.92. The paired t-test value is 2.02, and the p-value is less than 0.05. Therefore, we reject the null hypothesis.

This suggests that female twins report significantly more physical symptoms than male twins. In the Discussion section, the authors state that women report more physical symptoms than men, supporting behavioral theories of illness reporting. Thus, the statistical results support the authors’ interpretation.

## Full review

*Based on the elements provided above, write a review of this article highlighting its strengths and weaknesses in data analysis. Include specific recommendations for improvement where relevant. Your review should be written as a narrative (not bullet points) for an audience of public health researchers, in fewer than 200 words (include a word count).*

This study uses a population-based sample of unlike-sex twins to examine gender differences in health while controlling for genetic and early environmental factors. The matched-pair design is a major strength, as it reduces confounding and improves internal validity. The use of paired t-tests is appropriate for comparing outcomes between twins.

However, the authors provide limited information regarding the verification of statistical assumptions. No diagnostic plots or tests are presented to assess the normality of paired differences or identify outliers. In addition, the study focuses mainly on p-values and does not consistently report confidence intervals or effect sizes, limiting the interpretation of practical significance.

Future research should include graphical and statistical assessments of assumptions and report confidence intervals for all estimates. Sensitivity analyses using non-parametric methods would further strengthen the conclusions. Overall, the study makes an important contribution to understanding gender differences in health, but greater transparency in statistical reporting would improve its rigor.

# Data analysis

## Introduction

*Based on what you learned from the paper we just read, we want to further investigate potential drivers of differences in cardiovascular health across gender. Here, we want to assess if body mass index (BMI) is associated with gender in young adults.*

*This part of the homework will focus on the analysis of data from The National Health and Nutrition Examination Survey (NHANES), run by the CDC, collects information on the health and nutrition of people in the US through interviews, physical exams, and laboratory tests. It is widely used to monitor health trends and inform public health policy. The complete data set and accompanying information are available at [cdc.gov/nchs/nhanes](https://www.cdc.gov/nchs/nhanes/index.html).*

*For this assignment, you will work with a simplified subset of NHANES data set (same as Homework #01.) that includes a small number of variables and observations. This smaller data set is designed to let you practice data analysis.*

## Loading Data

```{r, results='hide'}

data <- read.csv("0219—CDC_NCHS_subset (1).csv")

head(data)
str(data)


```

## Visualizing and Summarizing Data

*Plot the distribution of BMI in males and females.*

```{r, results='hide'}

library(ggplot2)

ggplot(data, aes(x = BMI, fill = Gender)) +
  geom_histogram(alpha = 0.6, bins = 30, position = "identity") +
  facet_wrap(~Gender) +
  theme_minimal() +
  labs(title = "BMI Distribution by Gender",
       x = "BMI",
       y = "Count")


```

*Discuss BMI distribution.*

The BMI distributions for males and females are approximately bell-shaped with slight right skewness. Both groups show substantial overlap, although males tend to have a slightly higher average BMI. Some extreme values are present in both distributions.

*Do you expect to detect a significant association between gender and BMI based on the visual exploration of the data? Why?*

Based on the visual overlap between the two distributions, the difference in BMI between males and females appears modest. Therefore, a statistically significant association may be detected if the sample size is large, but the practical difference is likely small.

*Calculate the most appropriate summary statistics, stratified by gender.*

```{r, results='hide'}

library(dplyr)

data %>%
  group_by(Gender) %>%
  summarise(
    n = n(),
    mean_BMI = mean(BMI, na.rm = TRUE),
    sd_BMI = sd(BMI, na.rm = TRUE),
    median_BMI = median(BMI, na.rm = TRUE),
    IQR_BMI = IQR(BMI, na.rm = TRUE)
  )


```

The summary statistics show that the sample sizes for males and females are similar. The mean and median BMI are slightly higher among males than females, while the standard deviations and interquartile ranges are comparable between the two groups. This suggests that BMI variability is similar across genders, although males tend to have a somewhat higher central tendency.

*Run the appropriate statistical test to compare BMI across gender. Include all the steps, from stating the question to explaining the results in your written answer. You can re-use answers from questions above.*

```{r, results='hide'}

t.test(BMI ~ Gender, data = data)


```

The research question is whether BMI differs between males and females. An independent two-sample t-test was used because gender is categorical and BMI is continuous.

The null hypothesis states that mean BMI is equal between the two groups. Based on the t-test results, the p-value is less than 0.05, so we reject the null hypothesis. This indicates that males have a significantly higher average BMI than females in this sample.

# AI Use Disclosure Statement

I used ChatGPT to help clarify statistical concepts, improve wording, and assist with R programming. All analyses and interpretations were reviewed and verified by me.
